Designing and modeling a browser-based distributed evolutionary computation systems

### 2015-04-10 GECCO/modeling-volunteer-computing-gecco.tex
Cleaning up for next version

Another interesting result is that we have been able to obtain many
more different users. At the time of this writing we obtained 75
different IPs contributing CPU time to the experiment, giving an
average of around 4 per experiment. However, there is a big variation
in the number of CPUs. There is also a much bigger range in the
performance of these clients. In order to measure that we have
computed the time between two requests from the same IP and plotted it
as an histogram in Figure \ref{fig:intervals}.

@@

In this paper we have presented our experience on using browser-based
computing applied to the design of evolutionary algorithms. In the
spirit of Open Science, we have released all material related to the
experiment, including this paper. Experiments have been performed at
several times of the day and announced in Twitter or during a live
presentation about distributed computation using the browser.

What we conclude from these computers is that the nature of the
performance, which is due to the number of persons deciding to
participate in the experiment, is not totally random. There are at
least lower bounds we can count on: several computers (as few as six)
can almost always be relied on, and in some cases up to 30 can
participate in a single experiment. However, it is not clear how all
these computers contribute to the overall performance in terms of
time, although all experiments were performed until a solution was
found. Most computers participating in the experiment are in the same
performance tier, although a small percentage of them (which will be
around 25\%) are quite slow.

The evolutionary algorithm programmed in this way can easily
accommodate all kinds of browsers. However, we are using an homogeneous
configuration for all of them, which might not be the best from an
algorithmic point of view; a random parameter setting like the one
used by \cite{LNCS86720702} could offer better results, since we do
not know in advance what is the performance of the nodes. More
systematic experimentation is also needed, specially using different
kind of problems, including more complex problems in which the fitness
function is {\em heavier} and takes longer to be processed. However,
the main intention of this paper, which was to prove that distributed
evolutionary computation could be done efficiently in a volunteer/free
environment, has been sufficiently proved.

There are many issues involved in using these resources: from adapting
current algorithms so that they match this environment 
to check which EA configuration works the best in it, or creating a
framework that can use it easily. But the main
challenge is that asking people to contribute resources implies that
you are opening your science to society and you have to give something
in return: you have to adopt a set of best practices that have come to
be known as Open Science, because ``Give, and it shall be given unto
you'', you will get as much back from society as you give to it
opening your science and explaining it to the public. This, among
other things, means that popularity will become directly performance
of the metacomputer you create by attracting more users. And this
interplay between the social network itself (popularity in Twitter,
people attending to the conference and interested in following the
slides in their computer) is another, very interesting, venue to
explore in the future.

>>>>

Another interesting result is that we have been able to obtain many
more different users. At the time of this writing we obtained 75
different IPs contributing CPU time to the experiment, giving an
average of around 4 per experiment. However, there is a big variation
in the number of CPUs. There is also a much bigger range in the
performance of these clients. In order to measure that we have
computed the time between two requests from the same IP and plotted it
as an histogram in Figure \ref{fig:intervals}. In that histogram we
can see that there is an almost uniform distribution of intervals,
from the smallest taking a few milliseconds to the biggest with almost
1 second; the setup of the experiment allows it to be accessible from
a very wide range of different IPs. Even within a single IP the
variation can be relatively high, which is shown in a boxplot in
Figure \ref{fig:intervalIP}, sorted by median time. The time between
requests to the server, that is, the time to process 100 generations
of 128 individuals of an evolutionary algorithm to optimize the Trap
function has also very wide range, but even within the same IP,
inter-quartile differences of several hundred milliseconds are not
uncommon. Median for all IPs is 485 milliseconds and average is quite
close, at 496.8 ms.

Some interesting conclusions can be drawn from this data. The first is
not new: we need frameworks able to work with asynchronous algorithms
to take advantage of all the processing power that is available in an
ad-hoc, ephemeral, setup such as the one available in evolutionary
algorithms. The second is that there is going to be CPU available from
volunteers even when no active announcement is made, just from people
that come from a link in a social media page or the Twitter
timeline. However, an announcement brings immediate attention to the
page which will, obviously, depend on the popularity of the social
media and the person announcing it. How this dependency behaves is
still an open issue. Finally it is also an interesting conclusion
that, given enough time and free computing power, we can solve a
difficult optimization problem. Eventually.

@@

In this paper we have presented our experience on using browser-based
computing applied to the design of evolutionary algorithms and an
analysis of user behavior focused on the type of device, represented
by its performance, they will be using to access the URL that holds
the experiment. In the
spirit of Open Science, we have released all material related to the
experiment, including this paper. Experiments have been performed at
several times of the day and announced in Twitter, LinkedIn  or other
social media.

From the last experiment published we concluded that the nature of the
performance, which is due to the number of persons deciding to
participate in the experiment, is not totally random. There are at
least lower bounds we can count on: several computers (as few as six)
can almost always be relied on, and in some cases up to 30 can
participate in a single experiment. However, it is not clear how all
these computers contribute to the overall performance in terms of
time, although all experiments were performed until a solution was
found. In the last experiment, performed mainly in front of a live
audience using lapto computers, most of the participants in the experiment were in the same
performance tier, although a small percentage of them (which will be
around 25\%) are quite slow. In this experiment the conclusion is
different: uniform distribution of performance from quite slow to
blazing fast, with approximately 50\% slower than the average and,
conversely, 50\% faster. Our algorithm should, then, cater to a very
wide performance range so some methodology for self-adaptation should
be used on the evolutionary algorithm running in the browser. A more
precise model of the user is needed also in order to properly design
the algorithm, carrying it further away from the canonical
evolutionary algorithm we are using right now, which can only take
performance so far.

More
systematic experimentation is also needed, specially using different
kind of problems, including more complex ones in which the fitness
function is {\em heavier} and takes longer to be processed. However,
the main intention of this paper, which was to provide some initial
measures on how the users of an volunteer evolutionary algorithm
experiment would behave, has been reached.

There are many issues involved in using these resources: from adapting
current algorithms so that they match this environment 
to check which EA configuration works the best in it, or creating a
framework that can use it easily. But the main
challenge is that asking people to contribute resources implies that
you are opening your science to society and you have to give something
in return: you have to adopt a set of best practices that have come to
be known as Open Science, because ``Give, and it shall be given unto
you'', you will get as much back from society as you give to it
opening your science and explaining it to the public. In this paper,
we have shown some algorithmic visualization so that at least it is
visually attractive; in the previous experiment it is a
presentation. But there are many other possible ways of doing it,
including measuring the effect of a stealth algorithm which does not
shown anything at all. Some intriguing effects on how the experiment
is announced have also been noticed. While an announcement in English
attracted no new user, another in Spanish did attract immediately
several of them. Due to the popularity of WhatsApp groups, we would
like to test what happens with them, and how many will be visiting the
URL as a percentage of the users compared with the usage rate of
Twitter or Facebook.

It also remains to be done a comparison of single-desktop computer
experiment compared with this parallel setup. It is quite clear that,
due to overhead, several volunteers will be needed to match the speed
of a single several-core desktop computer. This will be explored in
future work.
